---
title: "R workshop"
author: "Jackie Maud"
date: "21 January 2019"
output: html_document
---


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load any packages here
library(tidyverse)


#load any data files here
```

# Day 2

## Tidy Data in spreadsheets


The functions for tidying data are:

:: tells which package to use function from

*tidyr::spread()
*tidyr::gather()


### Spread

3 principal arguments

1. the data
2. the _key_ column variable will become the new column names
3. the _value_ column variable whch will fill the new column variables

Use surveys dataset

Make from LONG to WIDE (spread)


```{r}
surveys<-read_csv(here::here("read_data", "surveys.csv"))

library(tidyverse)

#create a wide data format of surveys using spread

#first create a summary

surveys_gw <- surveys %>%
    drop_na(weight) %>% 
      group_by(species_id) %>% 
  summarize(mean_weight = mean(weight))


str(surveys_gw)

wide_surveys_gw <-surveys_gw %>% 
  spread(key = species_id, value = mean_weight)


```

### gather()


# Now back to long data from wide

gather takes 4 arguments

1. data
2. _key_
3. _value_
4. names of columns we use to fill the key variable (or drop)

```{r}
long_surveys_gw <-wide_surveys_gw %>% 
  gather(key = species_id, value = mean_weight)
```

## Sending Tidy Data

### Changelog

* Update your change log with changes to raw data/project


### Data dictionary

*Create to define our variables


```{r}
tidy_gsi <- read_csv(here::here("write_data", "tidy_gsi.csv"))
view(tidy_gsi)

```


