---
title: "R workshop"
author: "Jackie Maud"
date: "21 January 2019"
output: html_document
---


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load any packages here
library(tidyverse)


#load any data files here
```

# Day 2

## Tidy Data in spreadsheets


The functions for tidying data are:

:: tells which package to use function from

*tidyr::spread()
*tidyr::gather()


### Spread

3 principal arguments

1. the data
2. the _key_ column variable will become the new column names
3. the _value_ column variable whch will fill the new column variables

Use surveys dataset

Make from LONG to WIDE (spread)


```{r}
surveys<-read_csv(here::here("read_data", "surveys.csv"))

library(tidyverse)

#create a wide data format of surveys using spread

#first create a summary

surveys_gw <- surveys %>%
    drop_na(weight) %>% 
      group_by(species_id) %>% 
  summarize(mean_weight = mean(weight))


str(surveys_gw)

wide_surveys_gw <-surveys_gw %>% 
  spread(key = species_id, value = mean_weight)


```

### gather()


# Now back to long data from wide

gather takes 4 arguments

1. data
2. _key_
3. _value_
4. names of columns we use to fill the key variable (or drop)

```{r}
long_surveys_gw <-wide_surveys_gw %>% 
  gather(key = species_id, value = mean_weight)
```

## Sending Tidy Data

### Changelog

* Update your change log with changes to raw data/project


### Data dictionary

*Create to define our variables


```{r}
tidy_gsi <- read_csv(here::here("read_data", "tidy_gsi.csv"))
view(tidy_gsi)

```
# Analysing data

## Importing from Hakai Data Portal

Switched to data_wrangling script to import data into our read_data file

Chl_a, fish and sockeye stock ID data

```{r}
library(here)

fish <-read_csv(here("read_data", "fish.csv"))
                
chla <- read_csv(here("read_data", "chla.csv"))

tidy_gsi <- read_csv(here("read_data", "tidy_gsi.csv"))

fish %>% 
  count(species)
                
fish_d09 <- fish %>% 
  filter(site_id == "D09") %>% 
  select(hakai_id, jsp_survey_id, seine_id, date, species, site_id, fork_length, weight) %>% 
  mutate(k = (10^5 * weight) / fork_length^3) %>% 
  drop_na(k)
```

### Annoying things that will get you

## Factors

```{r}
str(fish_d09)

#or

class(fish_d09$species)


#coerce a column to be a factor

fish_d09$species <- factor(fish_d09$species)

levels(fish_d09$species)

```

If you have factors that are numbers, don't try to do maths with these Under the hood R will treat your factor levels as numbers.


## Dates

read_csv(): treats ISO date standards (yyyy-mm-dd) as a DATE object
read.csv(): treats them as characters - not ideal

Lubridate package - to help with dates

```{r}
library(lubridate)

# Extract data components (day, month, year, Julian day/yday)

fish_d09 <- fish_d09 %>% 
  mutate(year = year(date),
         month = month(date),
         week = week(date),
         yday = yday(date))

# to change format of column to DATE

#fish_d09$date <- as.date(fish_d09$date)
```

# Can do maths with lubridate

* periods
*intervals
*durations

##Joining data

Data:

*chla
*tidy_gsi
*fish_d09

```{r}

# see dplyr cheatsheet for help on this

left_join(fish_d09, tidy_gsi, by = "hakai_id") 

right_join(fish_d09, tidy_gsi, by = "hakai_id")

inner_join(fish_d09, tidy_gsi, by = "hakai_id")  #rows that have all required data

anti_join(fish_d09, tidy_gsi, by = "hakai_id")   #rows that DON'T have a match, i.e. NAs

#view()  displays results for last function (if haven't created it as new df)

```

##ggplot2

To build a ggplot:

ggplot(data = DATA, mapping = aes(MAPPINGS))
  + GEOM_FUNCTION()
  
  Example:
  
#ggplot(data = surveys, mapping = aes(species, weight)) +
  #+geom_point()
  
```{r}
ggplot(fish_d09, aes()) +
  geom_point(aes(x = yday, y = fork_length))+
  geom_smooth(aes(x = yday, y = fork_length), model = lm)+
  theme_classic()+
  facet_grid(species~.)    #separates data by specified variable

#or facet_wrap(species~.)

ggplot(fish_d09, aes()) +
  geom_point(aes(x = yday, y = fork_length))+
  geom_smooth(aes(x = yday, y = fork_length), model = lm)+
  theme_classic()+
  facet_wrap(species~.)  

# check out different bg themes, _bw (gridlines) and _classic (no gridlines)

```
  
  Cookbook for R http://www.cookbook-r.com/
  
  
  How to adjust legends, axes, etc.
  
  Geom list:
  
  tidyverse.org/reference
  
  https://dplyr.tidyverse.org/reference/index.html
  
  
  
  ## Playtime
  
```{r}


  
  
chla_date <- chla %>% 
    group_by(date) %>% 
    summarize(mean_chla = mean(chla, na.rm = TRUE))

chla_date <- chla_date %>% 
  mutate(year = year(date),
         month = month(date),
         week = week(date),
         yday = yday(date))
    

ggplot(chla_date, aes()) +
  geom_point(aes(x = yday, y = mean_chla))+
    theme_classic()+
    facet_grid(year~.)

ggplot(chla_date, aes()) +
  geom_point(aes(x = yday, y = mean_chla))+
    theme_classic()+
    facet_wrap(year~.)

  
  
           
  
  
  

```